Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-08 manuscript.tex
Update on Overleaf.

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008,villette_internally_2015}. Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.

In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019} reproduced below). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.

@@

@@

Let's now focus on distances defined based on a generative model formed by using precise spiking motifs. A review of the methods for neuronal response latency estimation exists and includes Bayesian binning~\citep{levakova_review_2015}. More broadly, unitary event analysis can be performed by a statistical model of coincidence detection~\citep{grun_unitary_2002-1}. This was extensively used in detecting above chance significant synchronous patterns~\citep{grun_unitary_2010}, in particular in recordings of pairs of neurons (see~\citep{riehle_spike_1997} for instance), and has been extended to non-stationary data~\citep{grun_unitary_2002}. A method to detect significant patterns of synchronous spiking in a subset of massively parallel spike trains in the presence of background activity can be defined using the statistical evaluation of synchronous spike patterns extracted by frequent item set mining~\citep{torre_statistical_2013}. By the same group, the SPADE, CAD or ASSET algorithms are methods for identification of spike patterns in massively parallel spike trains (the spiking activity of tens to hundred(s) of neurons recorded in parallel) by identifying fine temporal correlations in the$~\ms$ precision range~\citep{quaglio_methods_2018}. This was recently extended in~\citep{stella_3d-spade_2019} in order to find reoccurring spike patterns in parallel spike train data, and to determine their statistical significance. The extension improves the performance in the presence of patterns with different durations, as demonstrated by application to various synthetic data, such as surrogates generated to evaluate precisely timed higher-order spike correlations~\citep{stella_comparing_2022}.

% TODO: summarize ``Understanding Auditory Spectro-Temporal Receptive Fields and Their Changes with Input Statistics by Efficient Coding Principles'' - Spectro-temporal receptive fields (STRFs) have been widely used as linear approximations to the signal transform from sound spectrograms to neural responses along the auditory pathway. Their dependence on statistical attributes of the stimuli, such as sound intensity, is usually explained by nonlinear mechanisms and models. Here, we apply an efficient coding principle which has been successfully used to understand receptive fields in early stages of visual processing, in order to provide a computational understanding of the STRFs. According to this principle, STRFs result from an optimal tradeoff between maximizing the sensory information the brain receives, and minimizing the cost of the neural activities required to represent and transmit this information. Both terms depend on the statistical properties of the sensory inputs and the noise that corrupts them. The STRFs should therefore depend on the input power spectrum and the signal-to-noise ratio, which is assumed to increase with input intensity. We analytically derive the optimal STRFs when signal and noise are approximated as Gaussians. Under the constraint that they should be spectro-temporally local, the STRFs are predicted to adapt from being band-pass to low-pass filters as the input intensity reduces, or the input correlation becomes longer range in sound frequency or time. These predictions qualitatively match physiological observations. Our prediction as to how the STRFs should be determined by the input power spectrum could readily be tested, since this spectrum depends on the stimulus ensemble. The potentials and limitations of the efficient coding principle are discussed. ~\citep{zhao_understanding_2011}

%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
\subsection{Scaling up to very large scale data}

In~\citep{russo_cell_2017}, the authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which were applied to simultaneous electro-physiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity.

Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of \textit{in vivo} two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurophysics to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits.

One algorithm capable of achieving such a daunting task is the Rastermap algorithm~\citep{pachitariu_robustness_2018}. Basically, it re-arranges neurons in the raster plot based on similarity of activity and applies a deconvolution strategy based on a linear model. The model is \href{https://github.com/MouseLand/rastermap}{openly accessible} and has led to important discoveries. In~\citet{stringer_spontaneous_2019,stringer_high-dimensional_2019}, it was shown that a neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. 
Then, in~\citep{stringer_high-precision_2021}, the authors analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. Altogether, this provides evidence for the importance of such machine-learning-based tools to provide with breakthroughs in neuroscience.
% \href{https://nbviewer.org/github/MouseLand/rastermap/blob/master/tutorial/tutorial.ipynb}{Rastermap} re-arranges neurons in the raster plot based on similarity of activity

% \begin{itemize}
%  \item
%   https://www.janelia.org/lab/stringer-lab
% \item
%   described in
% \item
%   \href{https://github.com/MouseLand/rastermap}{rastermap}
% \item
%   deconvolution strategy
% \item
%   based on a linear model
% \end{itemize}
% \subsubsection{Stringer et al 2019, Nature~\citep{stringer_high-dimensional_2019}}\label{stringer-et-al-2019-nature-stringer2019nature}
% \begin{itemize}
%  \item
%   ``A neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. Here we analysed the dimensionality of the encoding of natural images by large populations of neurons in the visual cortex of awake mice.''
% \item
%   Data availability: All of the processed deconvolved calcium traces are available on \href{https://figshare.com/articles/Recordings_of_ten_thousand_neurons_in_visual_cortex_in_response_to_2_800_natural_images/6845348}{figshare}, together with the image stimuli.
% \item
%   Code availability: The code is available on \href{https://github.com/MouseLand/stringer-pachitariu-et-al-2018b}{GitHub}.
% \end{itemize}

% \subsubsection{Stringer et al 2019, Science~\citep{stringer_spontaneous_2019}}\label{stringer-et-al-2019-science-stringer2019science}
% \begin{itemize}
% \item
%   Stringer et al.~analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. The variability of neural responses to visual stimuli in the primary visual area is mainly related to arousal and reflects the encoding of latent behavioral states.
% \item
%   see also the work showing that you can encode very precise orientation information by using many neurons:~\citep{stringer_high-precision_2021}
% \end{itemize}

Another important algorithm is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}.
% \item
%   Detecting these temporal patterns represents a major methodological challenge.
% \item
%   Many approaches to this problem are supervised, that is, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity :
% \item
%   Nadasdy Z, Hirase H, Czurko A, Csicsvari J, Buzsaki G. Replay and time compression of recurring spike sequences in the hippocampus. J Neurosci. 1999;19(21):9497--507. pmid:10531452
% \item
%   Lee AK, Wilson MA. A combinatorial method for analyzing sequential firing patterns involving an arbitrary number of neurons based on relative time order. J Neurophysiol. 2004;92(4):2555--73. pmid:15212425
% \item
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631
% \item
In that method, the learning is unsupervised. It uses the prior that there is only one spike per pattern. 
% \item
Using a t-SNE projection with known labels validated that this clustering method can retrieve all patterns from the data.
% \item
%   data available @ https://doi.org/10.1371/journal.pcbi.1006283.s013
% \end{itemize}
The limits of this method is that it is block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.

%----------------------------%
\begin{figure}
\centering
%\includegraphics[width=7in,height=\textheight]{images/pcbi.1006283.g001.PNG_L.png}
\caption{Fig 1 of~\citep{grossberger_unsupervised_2018}: ``Simulated example illustrating the steps in SPOTDisClust. A) Structure of five ``ground-truth'' patterns (\ldots). For each pattern and each neuron, a random position was chosen for the activation pulse. B) Neuronal output is generated according to an inhomogeneous Poisson process, with rates dictated by the patterns in (A)." (© Authors under a \href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006283}{CC licence})}\label{fig:G2018-1}
\end{figure}
%----------------------------%

% \begin{itemize}
%  \item
%   ``However, SPOTDis has two principal weaknesses that we address here: (1) Its computational complexity, for comparing two time epochs, is O(N2), where N is the number of neurons. This becomes a major problem for computing an M dissimilarity matrix (for M time epochs) for thousands of neurons. (2) It relies exclusively on pairwise spike-timing relationships (i.e 2nd order correlations), because it does not solve the optimal transport problem for the entire spike pattern, but only for neuron pairs separately. Hence, it may not be sensitive to higher-order correlations in spiking
%   patterns.
%   Here, we develop a novel dissimilarity measure for multi-neuron spiking patterns called SpikeShip, which has linear computational complexity of O(N). We achieve this by (1) computing the minimum transport cost of spikes for each spike train separately, and (2) discounting a global translation term in the transport flow across neurons.''
%   https://doi.org/10.1101/2020.06.03.131573;
% \end{itemize}

\subsection{Extending the decoding at multiple time scales}

In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric test statistic for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.

% \subsubsection{Neural Variability and Sampling-Based Probabilistic Representations in the Visual Cortex~\citep{orban_neural_2016}}
% \begin{itemize}
%  \item
%   Stochastic sampling links perceptual uncertainty to neural response variability
% \item
%   Model accounts for independent changes in strength and variability of responses
% \item
%   Model predicts relationship between noise, signal, and spontaneous correlations
% \item
%   Stimulus statistics dependence of response statistics is explained
% \end{itemize}

To overcome the limits of models which require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.

% Currently, few interdisciplinary research projects take full advantage of event-based representations in computational and biological neuroscience. Nevertheless, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences (like synfire chains (Ikegaya et al., 2004), packets (Luczak et al., 2007) or hippocampal sequences (Pastalkova et al., 2008; Villette et al., 2015)  and on the role of such precise spike timing in downstream information transfer and coding (Branco et al., 2010; Buzsáki & Tingley, 2018; Luczak et al., 2015). In particular, one theoretical viewpoint considers synfire braids (Bienenstock, 1995), where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. In particular, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs (Izhikevich, 2006) which he defined as polychronous groups.

%Stereotyped sequences of neuronal activation have been particularly well described in the adult hippocampus and related to its function in mental travel in time and space (Buzsáki & Tingley, 2018). These sequences can be internally generated (Pastalkova et al., 2008; Villette et al., 2015) and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets (Malvache et al., 2016). Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure 2 of (Haimerl et al., 2019)). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.  
\begin{figure}
\centering
\includegraphics{figures/haimerl2019.jpg}
\caption{
%  Figure 2: An example of a polychronous group seen in cortical slices. In this study by (Haimerl et al., 2019), an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds.  
Mixed distance and duration representation in CA1. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse.}\label{fig:haimerl}
\end{figure}

% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.

In~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cortical songs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% what we say in section 2.3: Stereotyped sequences of neural activation have been particularly well described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019} below). 
We have seen previously that neurobiologists have described Stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019} below). 
Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called cortical songs. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out of about 900, and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spiking motifs in neuromorphic engineering}
\label{sec:neuromorphic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy2018toward}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical Von Neumann architectures is the localized memory of this new type of chips. It is materialized by the physical connections between the processing units that can store information, an example directly inspired by synaptic plasticity of such connection is the memristor (ref). Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges (zenke2021visualizing). Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware is described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field.

>>>>

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008}. Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.

In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and are formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. Thus, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019}). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.

@@

\includegraphics[width=\linewidth]{figures/haimerl2019.jpg}
\caption{An example of a polychronous group seen in cortical slices. In this study by~\citep{haimerl_internal_2019}, an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse.}\label{fig:haimerl}

@@

Let's now focus on distances defined based on a generative model formed by using precise spiking motifs. A review of the methods for neuronal response latency estimation exists and includes Bayesian binning~\citep{levakova_review_2015}. More broadly, unitary event analysis can be performed by a statistical model of coincidence detection~\citep{grun_unitary_2002-1}. This was extensively used in detecting above chance significant synchronous patterns~\citep{grun_unitary_2010}, in particular in recordings of pairs of neurons (see~\citep{riehle_spike_1997} for instance), and has been extended to non-stationary data~\citep{grun_unitary_2002}. A method to detect significant patterns of synchronous spiking in a subset of massively parallel spike trains in the presence of background activity can be defined using the statistical evaluation of synchronous spike patterns extracted by frequent item set mining~\citep{torre_statistical_2013}. By the same group, the SPADE, CAD or ASSET algorithms are methods for identification of spike patterns in massively parallel spike trains (the spiking activity of tens to hundred(s) of neurons recorded in parallel) by identifying fine temporal correlations in the$~\ms$ precision range~\citep{quaglio_methods_2018}. This was recently extended in~\citep{stella_3d-spade_2019} in order to find re-occurring patterns in parallel spike train data, and to determine their statistical significance. The extension improves the performance in the presence of patterns with different durations, as demonstrated by application to various synthetic data, such as surrogates generated to evaluate precisely timed higher-order spike correlations~\citep{stella_comparing_2022}.

Another important algorithm, called SPOTDisClust, is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}. In SPOTDisClust, the learning is unsupervised. It uses the prior that there is only one spike per pattern. Using a so-called t-SNE projection (that allows to project this high-dimensional representation to a lower-dimension map) validated that this clustering method can retrieve all patterns from the data (see Figure~\ref{fig:SPOTDisClust}). The limits of this method are that it is computationally complex, block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/pcbi.1006283.g001.png}%pcbi.1006283.g001.PNG_L.png}
\caption{Fig 1 of~\citep{grossberger_unsupervised_2018}, showing simulated example illustrating the steps in SPOTDisClust. A) Structure of five ``ground-truth'' patterns (\ldots). For each pattern and each neuron, a random position was chosen for the activation pulse. B) Neuronal output is generated according to an inhomogeneous Poisson process, with rates dictated by the patterns in (A). (© Authors under a \href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1006283}{CC licence})}\label{fig:SPOTDisClust}
\end{figure}
%----------------------------%
%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Scaling up to very large scale data}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of \textit{in vivo} two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurobiological knowledge to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits. As a consequence, there is a growing need to scale these methods to larger amounts of data.

There are multiple solutions which aim at tackling this problem. One algorithm capable of achieving such a daunting task is the Rastermap algorithm~\citep{pachitariu_robustness_2018}. Basically, it re-arranges neurons in the raster plot based on similarity of activity and applies a deconvolution strategy based on a linear model. The model is \href{https://github.com/MouseLand/rastermap}{openly accessible} and has led to important discoveries. In~\citet{stringer_spontaneous_2019,stringer_high-dimensional_2019} for instance, it was shown that a neuronal population encodes information most efficiently when its stimulus responses are high-dimensional and uncorrelated, and most robustly when they are lower-dimensional and correlated. Then, in~\citep{stringer_high-precision_2021}, the authors analyzed spontaneous neural firing, finding that neurons in the primary visual cortex encoded both visual information and motor activity related to facial movements. In~\citep{russo_cell_2017}, the authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which were applied to simultaneous electro-physiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity. Altogether, this provides evidence for the importance of such machine-learning-based tools to provide with breakthroughs in neuroscience.
%   Data availability: All of the processed deconvolved calcium traces are available on \href{https://figshare.com/articles/Recordings_of_ten_thousand_neurons_in_visual_cortex_in_response_to_2_800_natural_images/6845348}{figshare}, together with the image stimuli.

% \subsection{Extending the decoding at multiple time scales}
In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric test statistic for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.  To overcome the limits of models which either require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.
At a larger scale, in~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination. Building such models for predicting changes of timings based on context, such as using a so-called change point model for blocked-based experimental protocols~\citep{pasturel_humans_2020}, could therefore provide useful prior information to enhance the decoding from neural activity.

% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.
%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{cortical songs}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out of about 900, and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
Such precise structural information should therefore be included in decoding algorithm for a better understanding of neural activity.

@@

% \subsubsection{Neural Variability and Sampling-Based Probabilistic Representations in the Visual Cortex~\citep{orban_neural_2016}}
% \begin{itemize}
%  \item
%   Stochastic sampling links perceptual uncertainty to neural response variability
% \item
%   Model accounts for independent changes in strength and variability of responses
% \item
%   Model predicts relationship between noise, signal, and spontaneous correlations
% \item
%   Stimulus statistics dependence of response statistics is explained
% \end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Precise spiking motifs in neuromorphic engineering}
\label{sec:neuromorphic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Artificial intelligence has made huge advances in the past decades, now deep learning algorithms outperform humans at complex tasks like natural image recognition or abstract strategy board games. Machine learning algorithms still suffer from adversarial attacks or a lack of generalization capacity but their main weakness, compared to biological neural networks, rely on their poor energy efficiency. Neuromorphic engineering intends to mimic the neural bases of communication with a wide variety of technics, from strictly analog circuits to software-based neuromorphic systems, and to develop tools improving the capacities of current artificial intelligence (\cite{roy2018toward}). Because the reduced energy consumption of biological networks can be explained by the use of spikes and asynchronous responses to exchange information (ref), neuromorphic chips use this parallel and event-based representation to make computations. This is also the case for Dynamic Vision Sensors (DVS) which provide a stream of binary asynchronous events signaling detectable changes in luminance. These devices, also named ``silicon retinas'' (see Figure \ref{fig:silicon_retina}), show great improvements in terms of memory allocation for the recording of a visual scene (ref). Another important distinction with classical Von Neumann architectures is the localized memory of this new type of chips. It is materialized by the physical connections between the processing units that can store information, an example directly inspired by synaptic plasticity of such connection is the memristor (ref). Using these event-based computations as a building brick, neuromorphic engineering proposes new hardware designs perfect to simulate SNNs and use the full power of asynchronous computations observed in biological systems. Many connections can be drawn between neuromorphic engineering and computational neuroscience to aim at solving both research and technology challenges (zenke2021visualizing). Just as communication between neurons can be illustrated by a raster plot, information exchanges in neuromorphic hardware is described as event streams by an Address-Event Representation (AER). We will use indifferently \textit{events} or \textit{spikes} to describe the same binary and discrete processes and, in this section, we will extend the review of the use of precise spike timing to this emerging field.
